---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# diy.sem.plot
<!-- badges: start -->
<!-- badges: end -->

\pkg::{diy.sem.plot} allows you to manually plot fully customisable path diagrams for structural equation models (SEM) using the `diyPaths()` function. It avoids the inflexibility of automated SEM plotting tools and the tedium of drawing diagrams in external applications.

You map out where nodes go using simple coordinates and specify the attachment points on the perimeter of each node where paths begin and end. A range of fine-tuning options are included, facilitating the creation of a path diagram exactly as you envision it, entirely within R.

Currently diy.sem.plot only works for lavaan fitted models.

## Installation

Run the following to install the development version of the package.

``` {r, eval =FALSE}
install.packages("pak")
pak::pkg_install("snagy86/diy.sem.plot")
```

## Example

Below is a brief example of what the function is capable of. More examples and full explanation of work flow can be found in the package's vignette. 

```{r example1, fig.width = 100, fig.height = 10, out.width = "100%"}

# full SEM example, this model may not make theoretical sense.
library(diy.sem.plot)
library(lavaan)
library(ggplot2)

data(HolzingerSwineford1939, package = "lavaan")

# specify the model

sem_model <- '
   visual  =~ x1 + x2 + x3
   textual =~ x4 + x5 + x6
   speed   =~ x7 + x8 + x9

   speed ~ visual + textual
   visual ~~ textual
'

# fit the model

fit_sem <- sem(sem_model, data = HolzingerSwineford1939)

# specify the node position, this was done iteratively with show_grid to help with layout.
node_list <- list(
   # main latent variable structure
   node("visual", x = 1, y = 1, label = "Visual"),
   node("textual", x = 1, y = 3, label = "Textual"),
   node("speed", x = 4, y = 2, label = "Speed"),

   # observed variables that visual perception ability loads onto
   node("x1", x = 0, y = -0.5, label = "Visual\nPerception"),
   node("x2", x = 1, y = -0.5, label = "Cubes"),
   node("x3", x = 2, y = -0.5, label = "Lozenges"),

   # observed variables that textual ability loads onto
   node("x4", x = 0, y = 4.5, label = "Paragraph\nComprehension"),
   node("x5", x = 1, y = 4.5, label = "Sentence\nCompletion"),
   node("x6", x = 2, y = 4.5, label = "Word\nMeaning"),

   # observed variables that speeded cognitive processing loads onto
   node("x7", x = 6, y = 1.25, label = "Speeded\nAddition"),
   node("x8", x = 6, y = 2, label = "Speeded\nCounting"),
   node("x9", x = 6, y = 2.75, label = "Speeded\nDiscrimination")
)

# Specify the paths

path_list <- list(
  #Structural 
    path(from = "visual",  to = "speed", side_from = "right", side_to = "left"),
    #variance loop visual
     path(from = "visual",  to = "visual", variance_position = "top"),
   path(from = "textual", to = "speed", side_from = "right", side_to = "left"),
   path(from = "visual",  to = "textual", side_from = "left", side_to = "left", cov_curve = -0.6),
  
   #visual loadings
   path(from = "visual", to = "x1", side_from = "bottom", side_to = "top"), 
   path(from = "visual", to = "x2", side_from = "bottom", side_to = "top"),
   path(from = "visual", to = "x3", side_from = "bottom", side_to = "top"),

   #textual loadings
   path(from = "textual", to = "x4", side_from = "top", side_to = "bottom"),
   path(from = "textual", to = "x5", side_from = "top", side_to = "bottom"),
   path(from = "textual", to = "x6", side_from = "top", side_to = "bottom"),

   #speed loadings
   path(from = "speed",   to = "x7", side_from = "right", side_to = "left"),
   path(from = "speed",   to = "x8", side_from = "right", side_to = "left"),
   path(from = "speed",   to = "x9", side_from = "right", side_to = "left"),
   
  #latent variance/residualal
   path(from = "textual", to = "textual", variance_position = "bottom"),
   path(from = "visual", to = "visual", variance_position = "top"),
   path(from = "speed", to = "speed", variance_position = "top"),
  #Measurement Variances 
   path(from = "x1", to = "x1", variance_position = "bottom"),
   path(from = "x2", to = "x2", variance_position = "bottom"),
   path(from = "x3", to = "x3", variance_position = "bottom"),

   path(from = "x4", to = "x4", variance_position = "top"),
   path(from = "x5", to = "x5", variance_position = "top"),
   path(from = "x6", to = "x6", variance_position = "top"),

   path(from = "x7", to = "x7", variance_position = "right"),
   path(from = "x8", to = "x8", variance_position = "right"),
   path(from = "x9", to = "x9", variance_position = "right")
  
)

# creating the diagram

p <- diyPaths(
   fit = fit_sem,
   node_positions = node_list,
   path_positions = path_list,
   standardised = TRUE,
   sig_linetype = TRUE,
   observed_node_size_adjust = 0.5, 
   observed_node_text_size = 3.5,
   latent_node_text_size = 5,
   est_stars = TRUE,
   est_ci = TRUE,
   show_variances = TRUE,
   show_grid = TRUE,
   look_up_table = TRUE
)

print(p)
```

